### 2021-07-07 rad-icfp-2021/unf.tex
fixing unf

There are a few notable things in this syntax. 
Every primitive is indexed by a list of types from the source language, 
which corresponds to a context being carried. 
Every constant, unary, or binary operator from Source has a corresponding $n$-ary operator !op$_{T,n}$! in UNF.
Sequential composition is denoted by !;! and !e$_{1}$;e$_{2}$! means that !e$_{1}$! should be performed, and then !e$_{2}$!.
The array operators !map2! and !reduce! have extra indices that represent well-formed terms in Source.
The language has a second kind of sequential composition, which we call passing sequential compsition $\pcomp$.
In !e$_{1}$ $\pcomp$ e$_{2}$! the idea is first that !e$_{1}$! is computed, followed by !e$_{2}$!. 
However, !e$_{1}$! outputs an argument which is not used by !e$_{2}$! and is just passed by !e$_{2}$! 
along and returned with the output of !e$_{2}$!.
The language also contains a projection !proj! which forgets some of the elements, 
and a pairing operator !pair! which takes the last pair of elements and returns the pairing of these elements.

@@

The Target UNF is an extension of Source UNF. 
Its types are lists of types of Target. 
Higher-order types are needed for the continuation introduced by reverse-mode. 
In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
Lastly, Target UNF has an internal composition $\icomp$. 
!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!
It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
The grammar of Target UNF is given in Figure~\ref{fig:unf_source_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.

@@

We first give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}, which we call $\UNFSymbol$.
As discussed above, $\UNFSymbol$ sends a constant or an operator to a primitive in Source UNF. 
This term in Source UNF carries the context of the original term in Source.
$\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
Because of this, we we're trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
!e!$_{1} \pcomp$ !e!$_{2}$  is used to pass the result !e!$_{1}$ which is at the last position in the tuple through !e!$_{2}$.
It required when the coming operator represents a non-unary primitive such as $+$ or !pair!.
We want to preserve the invariant that $\UNFSymbol$(!e!) represents the term !e! which is also returning its context. 
Because !x! is not free in $\Gamma \vdash$ !let x=e$_1$in e$_2$!, we need to hide !x! after $\UNFSymbol$(!e$_2$!).
This explain the projection !proj! in $\UNFSymbol$ of a !let!. 
More generally, in !e!$_{1} \comp$ !e!$_{2}$ the return value of !e!$_{1}$ acts as a new variable for !e!$_{2}$, 
is not in the context of !e!$_{1}$, but will be returned by !e!$_{2}$.
This means we need to hide all these intermediate values to preserve the invariant,
and this explains the projection !proj! in the $\UNFSymbol$ of other terms.

TODO: we could merge everything of the form op;proj into just op by changing the type of op.

@@

Next, the transformation from Target UNF to Target is presented in Figure~\ref{fig:unf_to_target}. 
We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some kind of ANF. 
TODO: finish

>>>>

There are a few notable things in this syntax. 
Every primitive is indexed by a list of types from the source language, 
which corresponds to a context being carried. 
Every constant, unary, or binary operator from Source has a corresponding $n$-ary operator !op$_{T,n}$! in UNF.
Sequential composition is denoted by !;! and !e$_{1}$;e$_{2}$! means that !e$_{1}$! should be performed, and then !e$_{2}$!.
The array operators !map2! and !reduce! have extra indices that represent well-formed terms in Source.
The language also contains a projection !proj! which forgets some of the elements, 
and a pairing operator !pair! which takes the last pair of elements and returns the pairing of these elements.

@@

The Target UNF is an extension of Source UNF. 
Its types are lists of types of Target. 
Higher-order types are needed for the continuation introduced by reverse-mode. 
In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
Given a type !T$\defeq$ [A$_1$,$\ldots$,A$_n$]!, we write \underline{T} for the type !A$_1$ $\times\ldots\times$ A$_n$! of Target. 
Target UNF has an internal composition $\icomp$. 
!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!.
It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
The grammar of Target UNF is given in Figure~\ref{fig:unf_source_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.

@@

We first give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}, which we call $\UNFSymbol$.
As discussed above, $\UNFSymbol$ sends a constant or an operator to a primitive in Source UNF. 
This term in Source UNF carries the context of the original term in Source.
$\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
Because of this, we we're trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
When we need to pass the result of type !A! of !e$_{1}$! through !e$_{2}$! which does not take it as an input, we can transform !e$_{2}$! to accept and pass it.
It's the equivalent in UNF of a weakening.
This weakening is required when we transform a non-unary operator such as $+$ or !pair!, as we transform the arguments in order and we need to keep all of their results.
We write this new term $(\widetilde{e_{2}})_{A}$ and it defined by induction on !e$_{2}$! as follows. 
We assume given a type !A! that needs to be passed, and drop the index !A!.

\begin{tabular}{r c l}
    $\widetilde{var_{T;i}}$  &=& $var_{T,A;i}$ \\
    $\widetilde{op_{T;n}}$  &=& $op_{T,A;n}$ \\§
    $\widetilde{pair_{T;B\times C}}$  &=& $pair_{T,A;B\times C}$ \\
    $\widetilde{proj_{T_1;T_2;T_3}}$  &=& $proj_{T_1;T_2;T_3,A}$ \\
    $\widetilde{e_1 \comp e_2}$  &=& $\widetilde{e_1}\comp\widetilde{e_2}$ \\
    $\widetilde{map2_{T;x,y.e}}$  &=& $map2_{T,A;x,y.e}$ \\
    $\widetilde{reduce_{T;x,y.e_1;e_2}}$ &=& $reduce_{T,A;x,y.e_1;e_2}$ \\
\end{tabular}

We want to preserve the invariant that $\UNFSymbol$(!e!) represents the term !e! which is also returning its context. 
Because !x! is not free in $\Gamma \vdash$ !let x=e$_1$in e$_2$!, we need to hide !x! after $\UNFSymbol$(!e$_2$!).
This explain the projection !proj! in $\UNFSymbol$ of a !let!. 
More generally, in !e!$_{1} \comp$ !e!$_{2}$ the return value of !e!$_{1}$ acts as a new variable for !e!$_{2}$, 
is not in the context of !e!$_{1}$, but will be returned by !e!$_{2}$.
This means we need to hide all these intermediate values to preserve the invariant,
and this explains the projection !proj! in the $\UNFSymbol$ of other terms.

\begin{lemma}[Well typedness $\UNFSymbol$]
    If !$\Gamma$=x$_1$:A$_{1}$,$\ldots$,x$_n$:A$_{n}$ $\vdash$ e:B! then
   !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ $\UNFSymbol$(e): A$_{1}$,$\ldots$,A$_{n}$,B!.
\end{lemma}

% TODO: we could merge everything of the form op;proj into just op by changing the type of op.

@@

Next, the transformation from Target UNF to Target is presented in Figure~\ref{fig:unf_to_target}. 
We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of ANF transformation. 
$\UNFSymbol^{-1}$ does not preserve a typing property as nicely as $\UNFSymbol$, 
because it treats primitives with and without a $J$ differently.

Target UNF does not have variables. 
A context type !T=[A$_1$,$\ldots$,A$_n$]! is transformed to a context !$\Gamma$=x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$!.
We use this convention when definiting $\UNFSymbol^{-1}$. 
For operators like !op$_{\Gamma;m}$!, we extend the notation above by saying that the canonical variables are 
!x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$,x$_{n+1}$:$\reals$,$\ldots$,x$_{n+m}$:$\reals$!.
